---
title: "Bulk-RNA Seq of small intestinal organoids from young and aged mice"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tximportData)
library(ggrepel)
library(DESeq2)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(cowplot)

```


## B110 Theme

```{r theme, include=FALSE}

theme_b110 <- function(){
  theme_classic() +
  theme(
    axis.text = element_text(size = 10), 
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12,hjust = 0.5,face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "bottom"
    )
  
  theme_cowplot()
}

```

## B110 Colors

```{r colors}

b110_grey   = '#808080'
b110_grey_light   = '#909090'
b110_transparent_black = alpha('#000000',0.5)
google_red = '#dd4b39'
google_green = '#0F9D58'
google_yellow = '#F4B400'
google_blue = '#4285F4'
maja_red = alpha('#E64B35CC',0.8)
maja_blue = alpha('#4DBBD5CC',0.8)
maja_young = '#91D1C2CC'
maja_youngD = '#00A087CC'
maja_aged = '#8491B4CC'
maja_agedD = '#3C5488CC'

```

## Read in the meta data and exclude Wnt data

```{r meta_data}

raw_meta_data1 <- read_delim("../raw_data/15607_meta.tsv",delim = "\t")

raw_meta_data1 %<>% 
  filter(grepl("R1",FASTQ_FILE)) %>% 
  mutate(run = gsub(pattern = "(.+)_.*.fastq.gz",replacement = "\\1",x = FASTQ_FILE)) %>% 
  dplyr::rename(sample = SAMPLE_ID) %>% mutate(batch = "b15607")


raw_meta_data2 <- read_delim("../raw_data/16694_meta.tsv",delim = "\t")

raw_meta_data2 %<>% 
  filter(grepl("R1",FASTQ_FILE)) %>% 
  mutate(run = gsub(pattern = "(.+)_.*.fastq.gz",replacement = "\\1",x = FASTQ_FILE)) %>% 
  dplyr::rename(sample = SAMPLE_ID) %>% mutate(batch = "b16694")

raw_meta_data <- bind_rows(list(raw_meta_data1,raw_meta_data2))

rm(raw_meta_data1,raw_meta_data2)

raw_meta_data <- raw_meta_data %>% 
  filter(!grepl("Wnt", sample))

```

#import the data from star

```{r,warning=F,message=F}

star_data <- read_delim("../raw_data/star_merged_gene_counts.tsv",delim = "\t") %>% 
  select(Geneid, gene_name, starts_with(raw_meta_data$run))

colnames(star_data) <- gsub(pattern = "(\\S+?)Aligned.sortedByCoord.out.bam",replacement = "\\1",x = colnames(star_data))

star_data_id2name <- star_data %>% select(-starts_with("AS"))

star_data %<>% select(-gene_name) %>% column_to_rownames("Geneid") %>% as.data.frame()
```
  
#prepare the DESeq2 data object

```{r,warning=F,message=F}
require("DESeq2")

sampleTable<-
  raw_meta_data %>% 
  dplyr::select(sample,run,batch) %>% 
  mutate(Age=factor(if_else(grepl("Aged|-A",sample),"Aged","Young"))) %>%
  as.data.frame()

rownames(sampleTable)<-sampleTable$run



```

#simple prefiltering

```{r,warning=F,message=F}
sampleTable <- sampleTable[colnames(star_data),]
dds_star <- DESeqDataSetFromMatrix(star_data, sampleTable, ~batch+Age)

dds <- dds_star

nrow(dds)

#excluding genes that were not matched at all
dds <- dds[ rowSums(counts(dds)) > 1, ]
nrow(dds)

#remaining 24674

```

#stabilize variance for low readcount genes

```{r,warning=F,message=F}

vsd <- vst(dds, blind = FALSE)

```

# lets visualize the data with the batch effect removed

```{r}
require(patchwork)

vsd_batchrm <- vsd
assay(vsd_batchrm) <- limma::removeBatchEffect(assay(vsd_batchrm), vsd_batchrm$batch)

b <- plotPCA(vsd, "Age") + ggtitle("PCA by Age") + scale_color_manual(values=c(maja_agedD,maja_young)) + theme_cowplot()

ggsave("../Graphics/PCA_bulk_exvivo.pdf",b, width = 10,height = 10,units ="cm")

b

```


# lets call differential genes

```{r,warning=F,message=F}

dds <- DESeq(dds)


```

# Aged vs young

```{r,warning=F,message=F}

resultsNames(dds)

res <- results(dds, contrast = c("Age","Aged","Young"))

summary(res)

#downregulated genes upon ageing

down_regulated_genes <- res %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name)  %>% 
  arrange(padj) %>% 
  filter(log2FoldChange<0)
  
  
write_delim(down_regulated_genes,"../Results/top_table_organoids_Aged_vs_Young_DOWN.txt",delim = "\t")


#upregulated genes upon ageing
  
up_regulated_genes <- res %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name)  %>% 
  arrange(padj) %>% 
  filter(log2FoldChange>0)

write_delim(up_regulated_genes,"../Results/top_table_organoids_Aged_vs_Young_UP.txt",delim = "\t")

#All genes

bulk_organoids_all_genes <- res %>%
  as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name)  %>% 
  arrange(padj)

write_delim(bulk_organoids_all_genes,"../Results/bulk_organoids_all_genes.txt",delim = "\t")

summary(res)


library(cowplot)

#Volcano plot Top15 DEGs

p1_data <-res  %>%
  as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name) %>% 
  mutate(signif=if_else(padj<=0.1,if_else(log2FoldChange>0,"up","down"),"none")) %>%
  mutate(signif=if_else(log2FoldChange>-0.5 & log2FoldChange < 0.5,"none", signif)) %>%
  tidyr::drop_na()

p1 = ggplot(p1_data,aes(x=log2FoldChange,y=-log10(pvalue),col=signif,name = gene_name)) +
    geom_point()+
    ggtitle("bulk_organoids_Volcano_Top15")+
    scale_color_manual(values = c(maja_blue,b110_grey,maja_red)) +
    geom_vline(xintercept = c(0.5, -0.5), lty=2) + geom_hline(yintercept = 2,lty=2) +
    geom_label_repel(data=p1_data %>% arrange(desc(abs(stat))) %>% head(15), aes(label=gene_name), min.segment.length = 0, col="black")+
    theme_cowplot()

ggsave("../Graphics/bulk_organoids_Volcano_Top15.pdf",p1,w = 12,h = 12,units = "cm")

p1

#Volcano Plot of GOIs

p2_data <-res  %>%
  as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name) %>% 
  mutate(signif=if_else(padj<=0.1,if_else(log2FoldChange>0,"up","down"),"none")) %>%
  mutate(signif=if_else(log2FoldChange>-0.5 & log2FoldChange < 0.5,"none", signif)) %>%
  tidyr::drop_na()  
  
  p2=ggplot(p2_data,aes(x=log2FoldChange,y=-log10(pvalue),col=signif,name = gene_name)) +
    geom_point()+
    ggtitle("bulk_organoids_Volcano_DEGs_GOI")+
    scale_color_manual(values = c(maja_blue,b110_grey,maja_red)) +
    geom_vline(xintercept = c(0.5, -0.5), lty=2) + geom_hline(yintercept = 2,lty=2)+
    geom_label_repel(data=p2_data %>% filter(gene_name %in% c("Ccr2","Ly6e","P2rx7", "Isg15","Oasl2", "Tnf","Oas3", "Ifit3", "Ifitm3", "Dhx58", "Tnfrsf9")), aes(label=gene_name), min.segment.length = 0, col="black") +
    theme_cowplot()
  
ggsave("../Graphics/bulk_organoids_Volcano_DEGs_GOI.pdf",p2,w = 12,h = 12,units = "cm")

p2 

```


# Count plots of GOIs 

```{r,warning=F,message=F}
require("ggbeeswarm")

LUT <- star_data_id2name %>% column_to_rownames("Geneid")

GOI <- "Dhx58" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
a <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

GOI <- "Ccr2" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
b <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

GOI <- "P2rx7" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
c <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

GOI <- "Oasl2" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
d <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

GOI <- "Tnf" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
e <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

GOI <- "Ciita" #rownames(res1)[which.min(res$padj)]
topGene <- LUT %>% dplyr::filter(gene_name == GOI) %>% rownames_to_column("ENBSEMBLID") %>% pull(ENBSEMBLID)
geneCounts <- plotCounts(dds, gene = topGene, intgroup = c("Age"), returnData = TRUE)
geneCounts <- geneCounts %>% mutate(Age = factor(Age, levels = c("Young", "Aged"), ordered = TRUE))
f <- ggplot(geneCounts, aes(x = Age, y = count,fill=Age)) +
  geom_boxplot() +
  geom_point(size = 1.5) +ggtitle(LUT[GOI,1]) +
  scale_fill_manual(values = c(maja_young,maja_aged)) +
  ggtitle(paste0(GOI)) +
  theme_cowplot()

g <- (a+b)/(c+d)/(e+f)

ggsave("../Graphics/bulk_exvivo_Boxplots_GOI.pdf",g,width =16 ,height = 16,units = "cm")

g



```

#Gene set enrichment analysis

#Interpretation

Gene set annotations are downloaded from MSigDB
https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp#H
29.05.2020: 14:42 CET

h.all : Hallmark gene sets

Citation for hallmark collection:

Liberzon A, Birger C, Thorvaldsdóttir H, Ghandi M, Mesirov JP, Tamayo P. The Molecular Signatures Database (MSigDB) hallmark gene set collection. Cell Syst. 2015 Dec 23;1(6):417-425. 
 

Signature enrichment

```{r}

require(fgsea)
require(msigdbr)
 
#first we get the recent gene set expression signatures
 
m_df <- msigdbr(species = "Mus musculus", category = "H")
head(m_df)
 
#format them for fgsea
 
m_list = m_df %>% split(x = .$gene_symbol, f = .$gs_name)

this_data <- res %>% as.data.frame() %>% 
  rownames_to_column(var = "Geneid") %>% 
  left_join(star_data_id2name)

seq_results <- this_data %>%
  select(stat=stat,Symbol=gene_name,Geneid) %>%
  column_to_rownames("Geneid") %>%
  na.omit() %>% 
  distinct() %>% 
  group_by(Symbol) %>% 
  summarize(stat=mean(stat)) %>%
  ungroup() %>%
  deframe()

#GSEA of hallmark sets

fgsea_hallmark <- fgsea(pathways=m_list, stats=seq_results, nperm=1000)

fgsea_hallmarktdy <- fgsea_hallmark %>%
  as_tibble() %>%
  arrange(desc(NES))

write_delim(fgsea_hallmarktdy,"../Results/fgsea_hallmarktdy_exvivo.txt",delim = "\t")

a = fgsea_hallmarktdy %>% 
  arrange(padj) %>%
  head(.,40) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA") + 
  theme_b110()+
  scale_fill_manual(values = c(b110_grey,google_blue))

ggsave("../Graphics/star_enriched_signatures_organoids.pdf",a, width = 10, height = 20)

a

```


#SessionInfo

```{r}
sessionInfo()
```










